---
title: "NZ Floral Volatilome Analysis"
author: "Flore Mas & Tom Moore"
date: last-modified
output-file: "2_voc_analysis.html"
format:
  html:
    toc: true
    toc-depth: 5
    toc-title: "Contents"
    theme: united
    highlight-style: tango
    fig-width: 9
    fig-height: 6
    code-tools: true
    code-fold: false
    code-overflow: scroll
    df-print: paged
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| results: false

# --- Packages: install if missing, then load ---
if (!requireNamespace("pak", quietly = TRUE)) install.packages("pak")
packages <- c(
  "tidyverse", "readxl", "here", "vegan", "janitor", "tidytext", "ggvenn",
  "reshape2", "remotes", "devtools", "ape", "ggtree", "caper", "phytools",
  "patchwork", "RColorBrewer"
)
missing_pkgs <- packages[!sapply(packages, requireNamespace, quietly = TRUE)]
if (length(missing_pkgs) > 0) {
  pak::pkg_install(missing_pkgs, ask = FALSE)
  cat("Installed:", paste(missing_pkgs, collapse = ", "), "\n")
}
invisible(sapply(packages, library, character.only = TRUE))
# geomorph (optional: requires X11/OpenGL on macOS)
if (requireNamespace("geomorph", quietly = TRUE)) {
  tryCatch(suppressPackageStartupMessages(library("geomorph", character.only = TRUE)), error = function(e) NULL)
}

# GitHub packages (ggvegan, V.PhyloMaker2)
if (!requireNamespace("ggvegan", quietly = TRUE)) remotes::install_github("gavinsimpson/ggvegan")
if (!requireNamespace("V.PhyloMaker2", quietly = TRUE)) devtools::install_github("jinyizju/V.PhyloMaker2")
library("ggvegan")
library("V.PhyloMaker2")
```

*The chemical diversity of floral volatilomes in native and exotic plants shaping Aotearoa-New Zealand's odourscapes, implications for plant-insect interactions*

Mas F.^1^, Manning L-A.^1^, Moore T.^1^, Cutting BT.^2^ and Evans LJ.^2^

^1^New Zealand Institute for Bioeconomy Science Limited, Lincoln 7608, New Zealand.

^2^New Zealand Institute for Bioeconomy Science Limited, Ruakura 3241, New Zealand.

# Overview

This report analyses floral volatile organic compounds (VOCs) from New Zealand plants. We use a cleaned dataset of compounds by sample, with plant species classified as **Natives**, **Crops**, or **Ornamentals**. The analysis covers: (1) chemical class distribution across categories, (2) diversity metrics (richness, Shannon/Hill, evenness) and their comparison by category, (3) overlap of compounds between categories (Venn), (4) seasonal phenology of VOC composition, and (5) supplementary ordination (NMDS), dispersion, and phylogeny. Statistical tests include chi-square (class vs category), ANOVA and post-hoc Wilcoxon (diversity), and Kruskal–Wallis (phenology); test outputs are in collapsible sections so you can show or hide them.

AI was used to format the analysis and this Quarto document.

::: {.callout-note title="Dataframe (click to show/hide)" collapse="true"}
## Dataframe

```{r}
#| label: data_cleaning
#| echo: false

# --- Load and clean raw VOC data ---
# Load raw VOC data and coerce to factors
df <- read_excel(here::here("Analysis/Data/voc_ldata.xlsx"), sheet = "LDATA_ALL_tous3")
df <- as.data.frame(unclass(df), stringsAsFactors = TRUE)

# Clean names and standardise factor labels
df <- df |> clean_names() |> as.data.frame(stringsAsFactors = TRUE) |>
  mutate(across(where(is.factor), ~ as.factor(str_to_sentence(as.character(.)))))

# Numeric columns
cols_numeric <- c("sum_area", "area", "tot_voc", "n_voc")
df[, cols_numeric] <- lapply(df[, cols_numeric], as.numeric)

# Sample-level ID and category (Natives / Crops / Ornamentals)
df <- df |> mutate(
  unique_id = as.factor(paste(as.character(sample), substr(as.character(location), 1, 3), sep = "-")),
  new_cat = case_when(
    native_exotic == "Native" ~ "Natives",
    native_exotic == "Exotic" & crop_wild == "Crop" ~ "Crops",
    native_exotic == "Exotic" & crop_wild == "Wild" ~ "Ornamentals",
    TRUE ~ NA_character_
  ),
  new_cat = as.factor(new_cat)
)

# Chemical class labels: rename and set order
df <- df |> mutate(
  new_class = fct_recode(new_class, "Fatty acid derivatives" = "Fatty_acid_derivatives"),
  new_sub_class = fct_recode(new_sub_class,
    "Benzenoids unknown" = "Benzenoids_unknown", "Fatty unknown" = "Fatty_unknown",
    "Fatty alcohols" = "Fatty_alcohols", "Fatty aldehydes" = "Fatty_aldehydes",
    "Fatty esters" = "Fatty_esters", "Phenolic acids" = "Phenolic_acids",
    "Irregular terpenes" = "Irregular_terpenes", "Terpenoids unknown" = "Terpenoids-unknown")
)
df$new_class <- factor(df$new_class, levels = c("Fatty acid derivatives", "Terpenoids", "Benzenoid", "NA"))
df$new_sub_class <- factor(df$new_sub_class,
  levels = c("Hydrocarbons", "Fatty esters", "Fatty aldehydes", "Fatty alcohols", "Fatty unknown", "Lactones",
    "Monoterpenes", "Sesquiterpenes", "Irregular terpenes", "Terpenoids unknown", "Benzenoids unknown",
    "Phenylpropanoids", "Phenolic acids", "Phenylethanoids", "NA"))

# Keep required columns and compute relative abundance per sample
df <- df |> dplyr::select(compound, unique_id, plant, type, new_cat, genus, family, order,
  location, region, new_class, new_sub_class, sum_area, perc_area, prop_area, smiles, inchikeys)
df <- df |> group_by(compound, unique_id) |> mutate(abundance = ifelse(is.na(sum_area), perc_area, sum_area)) |>
  ungroup() |> group_by(unique_id) |> mutate(sum_abund = sum(abundance)) |> ungroup() |>
  group_by(compound, unique_id) |> mutate(rel_abund = abundance / sum_abund) |> ungroup()

# Reclassify Trifolium/Lavandula as Ornamentals
df <- df |> mutate(
  new_cat = case_when(
    plant %in% c("Trifolium incarnatum", "Trifolium pratense", "Trifolium repens",
                 "Lavandula angustifolia", "Lavandula dentata") ~ "Ornamentals",
    TRUE ~ as.character(new_cat)
  ),
  new_cat = factor(new_cat)
)
```

**Dataframe structure** (`str(df)`):

```{r}
#| label: str_df
#| echo: false

# Show structure of main dataframe
str(df)
```
:::

::: {.callout-note title="Descriptive statistics (click to show/hide)" collapse="true"}
## Descriptive statistics

Summaries by sample, plant, category, and chemical class (shown below).

```{r}
#| label: descriptive_statistics
#| echo: false

# --- Descriptive summaries ---
# By sample (unique_id): compound count and total VOC
summary_unique_id <- df |> group_by(unique_id) |>
  summarise(unique_id_compound = n_distinct(compound), total_voc = sum(abundance))

# By plant: compound count, replicates, mean total VOC
summary_plant <- df |> group_by(unique_id) |>
  mutate(unique_id_compound = n_distinct(compound), total_voc = sum(abundance)) |>
  ungroup() |> group_by(plant) |>
  summarise(type = first(new_cat), count_compound = n_distinct(compound),
    n_replicates = n_distinct(unique_id), mean_total_voc = mean(total_voc), .groups = "drop")

# By category (new_cat): plants, compound stats, total VOC range
summary_categories <- df |> group_by(unique_id) |>
  mutate(unique_id_compound = n_distinct(compound), total_voc = sum(abundance)) |>
  ungroup() |> group_by(plant) |>
  mutate(count_compound = n_distinct(compound), nb_replicates = n_distinct(unique_id), mean_total_voc = mean(total_voc)) |>
  ungroup() |> group_by(new_cat) |>
  summarise(nb_plant = n_distinct(plant), mean_comp = mean(count_compound), median_voc = median(count_compound),
    min_comp = min(count_compound), max_comp = max(count_compound),
    mean_total_voc = mean(mean_total_voc), min_total_voc = min(mean_total_voc), max_total_voc = max(mean_total_voc))

# By chemical class: count and percent
summary_chem_class <- df |> group_by(new_class) |> summarise(count = n()) |> mutate(percent = 100 * count / sum(count))

# Display tables in report
head(summary_unique_id, 10)
summary_plant
summary_categories
summary_chem_class
```
:::

# Figures & hypothesis tests

## Figure 1
```{r}
#| label: fvoc_frequency
#| fig.width: 9
#| fig.height: 6

# Top 10 compounds by chemical class (Benzenoid, Fatty acid derivatives, Terpenoids)
df_ranked_class <- df |> group_by(new_class) |> count(compound, sort = TRUE) |>
  slice_max(n, n = 10, with_ties = FALSE) |> ungroup() |>
  filter(new_class != "NA") |> droplevels() |>
  mutate(new_class = fct_relevel(new_class, "Benzenoid", "Fatty acid derivatives", "Terpenoids"))

ggplot(df_ranked_class, aes(x = reorder(compound, n), y = n, fill = new_class)) +
  geom_bar(stat = "identity", alpha = 0.6) + coord_flip() +
  scale_fill_manual(name = "Compound Class", values = c("#A8A6A7FF", "#006A8EFF", "#B1283AFF")) +
  labs(x = "Floral Volatile Organic Compounds", y = "Count") +
  theme_classic() + theme(legend.position = "right")
ggsave(here::here("Plots", "Figure1.png"), width = 9, height = 6, dpi = 600)
```

## Figure 2
```{r}
#| label: chemical_class_distribution
#| fig.width: 9
#| fig.height: 3

# Chemical class proportions by category (donut-style pie charts)
df_pie <- df |> filter(new_class != "NA") |> droplevels() |>
  group_by(new_cat, new_class) |> count() |> group_by(new_cat) |>
  mutate(percent = (n / sum(n)) * 100) |> ungroup()
df_pie$new_class <- factor(df_pie$new_class, levels = c("Benzenoid", "Fatty acid derivatives", "Terpenoids"))

ggplot(df_pie, aes(x = "", y = percent, fill = new_class)) +
  geom_bar(stat = "identity", width = 1, alpha = 0.8, aes(color = new_cat), linewidth = 1.5) +
  coord_polar("y", start = 0, direction = -1) + facet_wrap(~new_cat, ncol = 3) +
  scale_fill_manual(name = "Compound Class", values = c("#A8A6A7FF", "#006A8EFF", "#B1283AFF")) +
  scale_color_manual(values = c("Crops" = "#DDCC77", "Natives" = "darkgreen", "Ornamentals" = "orange"), guide = "none") +
  theme_void() + theme(legend.position = "bottom", strip.text = element_text(size = 12, face = "bold"))
ggsave(here::here("Plots", "Figure2.png"), width = 9, height = 3, dpi = 600)
```

::: {.callout-note title="Chi-square test (chemical class vs category)" collapse="true"}
```{r}
#| label: chisq_fig2

contingency_table <- df |> filter(new_class != "NA") |> count(new_cat, new_class) |>
  pivot_wider(names_from = new_class, values_from = n, values_fill = 0) |> column_to_rownames("new_cat")
print(chisq.test(contingency_table))
```
:::

## Figure 3
```{r}
#| label: ven_diagram
#| fig.width: 9
#| fig.height: 6

# Venn: unique compounds per category (Crops, Natives, Ornamentals)
venn_data <- df |> group_by(new_cat) |> summarise(compounds = list(unique(compound))) |> deframe()
ggvenn(venn_data, fill_color = c("#DDCC77", "darkgreen", "orange"), fill_alpha = 0.5, stroke_size = 0.5, set_name_size = 5, text_size = 3)
ggsave(here::here("Plots", "Figure3.png"), width = 9, height = 6, dpi = 600)
```

## Figure 4
```{r}
#| label: top_10
#| fig.width: 9
#| fig.height: 3.5

# Top 10 VOCs per category
df_ranked <- df |> group_by(new_cat) |> count(compound, sort = TRUE) |>
  slice_max(n, n = 10, with_ties = FALSE) |> ungroup()

ggplot(df_ranked, aes(x = reorder_within(compound, n, new_cat), y = n, fill = new_cat)) +
  geom_bar(stat = "identity", alpha = 0.6) + coord_flip() + facet_wrap(~new_cat, ncol = 3, scales = "free") +
  scale_x_reordered() + scale_fill_manual(values = c("#DDCC77", "darkgreen", "orange")) +
  labs(x = "Top 10 Volatile Organic Compounds", y = "Count") +
  theme_classic() + theme(legend.position = "none", strip.text = element_text(size = 12, face = "bold"))
ggsave(here::here("Plots", "Figure4.png"), width = 9, height = 3.5, dpi = 600)
```

## Figure 5
```{r}
#| label: hills_diversity
#| fig.width: 9
#| fig.height: 3.5

# Wide format: one row per sample, columns = compounds (abundance)
df_wide <- df |> reshape2::dcast(unique_id + new_cat + location + region + family + order + plant ~ compound, value.var = "abundance", fun.aggregate = sum)
df_compounds <- df_wide |> dplyr::select(where(is.numeric))

# Disparity analysis (mean within- and between-group Euclidean distances)
dist_matrix <- dist(df_compounds, method = "euclidean")
disparity <- meandist(dist_matrix, df_wide$new_cat)

# Diversity: richness (q=0), Shannon/Hill (q=1), evenness (Pielou J)
richness <- specnumber(df_compounds)
shannon_div <- vegan::diversity(df_compounds, index = "shannon")
hill_q1 <- exp(shannon_div)
evenness <- shannon_div / log(richness)
diversity_df <- data.frame(
  unique_id = df_wide$unique_id, new_cat = df_wide$new_cat, family = df_wide$family,
  order = df_wide$order, plant = df_wide$plant, richness = richness, hill_q1 = hill_q1, evenness = evenness
)

# Boxplots (Richness, Evenness, Hill) by category
pal <- c("#DDCC77", "darkgreen", "orange")
p_richness <- ggplot(diversity_df, aes(x = new_cat, y = richness, fill = new_cat, color = new_cat)) +
  geom_jitter(alpha = 0.1, width = 0.3) + geom_boxplot(width = 0.3, alpha = 0.5, outlier.shape = NA) +
  scale_y_continuous(breaks = seq(0, max(diversity_df$richness), by = 10)) +
  scale_color_manual(values = pal) + scale_fill_manual(values = pal) +
  labs(y = "Richness", x = "") + theme_bw() +
  theme(legend.position = "none", axis.text = element_text(size = 10), axis.title = element_text(size = 11, face = "bold"))
p_evenness <- ggplot(diversity_df, aes(x = new_cat, y = evenness, fill = new_cat, color = new_cat)) +
  geom_jitter(alpha = 0.1, width = 0.3) + geom_boxplot(width = 0.3, alpha = 0.5, outlier.shape = NA) +
  scale_y_continuous(limits = c(0, 1.01), breaks = seq(0, 1, 0.2)) +
  scale_color_manual(values = pal) + scale_fill_manual(values = pal) +
  labs(y = "Evenness", x = "") + theme_bw() +
  theme(legend.position = "none", axis.text = element_text(size = 10), axis.title = element_text(size = 11, face = "bold"))
p_hill <- ggplot(diversity_df, aes(x = new_cat, y = hill_q1, fill = new_cat, color = new_cat)) +
  geom_jitter(alpha = 0.1, width = 0.3) + geom_boxplot(width = 0.3, alpha = 0.5, outlier.shape = NA) +
  scale_color_manual(values = pal) + scale_fill_manual(values = pal) +
  labs(y = "Hill Diversity", x = "") + theme_bw() +
  theme(legend.position = "none", axis.text = element_text(size = 10), axis.title = element_text(size = 11, face = "bold"))
p_richness + p_evenness + p_hill + plot_layout(ncol = 3) + plot_annotation(tag_levels = "A")
ggsave(here::here("Plots", "Figure5.png"), width = 9, height = 3.5, dpi = 600)
```

::: {.callout-note title="Figure 5: Disparity (meandist), MRPP, Kruskal-Wallis and post-hoc Wilcoxon (diversity by category)" collapse="true"}
```{r}
#| label: stats_fig5

# Disparity: mean within- and between-group distances
print(disparity)
summary(disparity)

# MRPP (Multi Response Permutation Procedure) on Euclidean distances
vegan::mrpp(dat = df_compounds, grouping = df_wide$new_cat, distance = "euclidean", permutations = 999)

# Diversity tests
kruskal.test(richness ~ new_cat, data = diversity_df)
kruskal.test(hill_q1 ~ new_cat, data = diversity_df)
kruskal.test(evenness ~ new_cat, data = diversity_df)
pairwise.wilcox.test(diversity_df$richness, diversity_df$new_cat, p.adjust.method = "BH")
pairwise.wilcox.test(diversity_df$hill_q1, diversity_df$new_cat, p.adjust.method = "BH")
pairwise.wilcox.test(diversity_df$evenness, diversity_df$new_cat, p.adjust.method = "BH")
```
:::

## Figure 6
```{r}
#| label: regional_crops
#| fig.width: 9
#| fig.height: 4.5

# Crops only: chemical class proportions by region
df_crops <- df |> filter(type %in% c("Arable", "Fruit"), !is.na(new_class)) |> droplevels()
df_pie <- df_crops |> group_by(region, new_class) |> count() |> group_by(region) |>
  mutate(percent = (n / sum(n)) * 100) |> ungroup()
df_pie <- df_pie |> mutate(
  region = fct_recode(region,
    "Tropical fruit" = "Central north island",
    "Pipfruit & Stonefruit" = "Lower north island",
    "Arable crops" = "Upper south island",
    "Summerfruit" = "Lower south island",
    "Natives" = "Aotearoa"),
  new_class = fct_relevel(new_class, "Benzenoid", "Fatty acid derivatives", "Terpenoids")
)

ggplot(df_pie, aes(x = "", y = percent, fill = new_class)) +
  geom_bar(stat = "identity", alpha = 0.8) + coord_polar("y", start = 0, direction = -1) +
  facet_wrap(~region) +
  scale_fill_manual(name = "Compound Class", values = c("#A8A6A7FF", "#006A8EFF", "#B1283AFF")) +
  theme_void() + theme(legend.position = "bottom", legend.title = element_blank(), strip.text = element_text(size = 10))
ggsave(here::here("Plots", "Figure6.png"), width = 9, height = 4.5, dpi = 600)
```

## Figure 7
```{r}
#| label: phenology
#| fig.width: 9
#| fig.height: 6.5

# Phenology: flowering season per plant; merge with VOC data
pheno <- read_excel(here::here("Analysis/Data/flower_phenology.xlsx"), sheet = "Sheet1")
pheno <- as.data.frame(unclass(pheno), stringsAsFactors = TRUE) |> clean_names() |>
  mutate(across(where(is.factor), ~ as.factor(str_to_sentence(as.character(.)))))
df_pheno <- df |> left_join(pheno, by = join_by(plant), relationship = "many-to-many") |> droplevels()
df_pheno <- df_pheno |> mutate(season = fct_recode(season, "Autumn" = "Fall")) |> droplevels()
df_pheno$season <- factor(df_pheno$season, levels = c("Spring", "Summer", "Autumn", "All-year"))

# Filter: exclude All-year and NA chemical class; mean rel_abund by category × class × season
df_pheno_filtered <- df_pheno |> filter(season != "All-year" & new_class != "NA") |> droplevels() |>
  mutate(new_class = fct_relevel(new_class, "Benzenoid", "Fatty acid derivatives", "Terpenoids"))
mean_df <- df_pheno_filtered |> group_by(new_cat, new_class, season) |>
  summarise(mean_abund = mean(rel_abund, na.rm = TRUE), .groups = "drop")
count_df <- df_pheno_filtered |> group_by(new_cat, season) |> summarise(nb_plant = n_distinct(plant), .groups = "drop")

# Panel A: mean relative abundance by season and chemical class (by category)
p1 <- ggplot(df_pheno_filtered, aes(x = season, y = rel_abund, colour = new_class, group = new_class)) +
  geom_line(data = mean_df, aes(x = season, y = mean_abund, colour = new_class, group = new_class), linewidth = 1, position = position_dodge(width = 0.1)) +
  geom_point(data = mean_df, aes(x = season, y = mean_abund, colour = new_class), position = position_dodge(width = 0.1), size = 2.5) +
  facet_grid(new_cat ~ .) +
  scale_colour_manual(name = "Chemical Class", values = c("Benzenoid" = "#A8A6A7FF", "Fatty acid derivatives" = "#006A8EFF", "Terpenoids" = "#B1283AFF")) +
  labs(title = "A) Seasonal patterns of chemical classes", y = "Mean relative abundance", x = "Season") +
  theme_classic() + theme(legend.position = "right", strip.text = element_text(size = 12, face = "bold"))
# Panel B: number of plants flowering per season
p2 <- ggplot(count_df, aes(x = season, y = nb_plant, fill = new_cat)) +
  geom_bar(stat = "identity", position = "dodge", width = 0.6, alpha = 0.8) +
  scale_fill_manual(name = "Category", values = c("#DDCC77", "darkgreen", "orange")) +
  labs(title = "B) Number of flowering plants per season", x = "Season", y = "Number of plants") +
  theme_classic() + theme(legend.position = "right", strip.text = element_text(size = 12, face = "bold"))
(p1 / p2) + plot_layout(heights = c(3, 1))
ggsave(here::here("Plots", "Figure7.png"), width = 9, height = 6.5, dpi = 600)
```

::: {.callout-note title="Figure 7: Kruskal–Wallis and post-hoc Wilcoxon (phenology by category × class)" collapse="true"}
```{r}
#| label: stats_phenology

for (cat in levels(df_pheno_filtered$new_cat)) {
  cat_data <- df_pheno_filtered |> filter(new_cat == cat)
  for (class in levels(cat_data$new_class)) {
    class_data <- cat_data |> filter(new_class == class)
    if (nrow(class_data) > 0 & length(unique(class_data$season)) > 1) {
      kw_test <- kruskal.test(rel_abund ~ season, data = class_data)
      cat(cat, "-", class, ": Kruskal-Wallis p =", format.pval(kw_test$p.value, digits = 3), "\n")
      if (kw_test$p.value < 0.05) print(pairwise.wilcox.test(class_data$rel_abund, class_data$season, p.adjust.method = "BH"))
    }
  }
}
```
:::

# Supplementary material

::: {.callout-note title="Table S1 (click to show/hide)" collapse="true"}
Summary of plant species collected by category (New Zealand native plants, and crops and ornamentals—both categories comprising plants exotic to New Zealand) with their number of unique floral volatile organic compounds (FVOCs).

```{r}
#| label: supp_table_category_family_plant
#| results: asis

supp_table <- df |> group_by(new_cat, family, plant) |> summarise(n_FVOC = n_distinct(compound), .groups = "drop") |> rename(Category = new_cat, Family = family, `Plant species` = plant, `Number of FVOCs` = n_FVOC)
cat("<style>.supp-table-full table { width: 100% !important; table-layout: fixed; } .supp-table-full td { word-wrap: break-word; overflow-wrap: break-word; font-size: 0.9em; }</style>")
cat("<div class='supp-table-full' style='font-size: 0.9em; width: 100%;'>")
cat(knitr::kable(supp_table, format = "html"))
cat("</div>")
```
:::

```{r}
#| label: supp_compound_summary
#| results: asis

compounds_by_class <- df |> distinct(compound, .keep_all = TRUE) |> count(new_class, name = "n_compounds") |> mutate(pct = 100 * n_compounds / sum(n_compounds))
n_total <- sum(compounds_by_class$n_compounds)
cat("**Total different compounds detected:** ", n_total, "\n\n")
cat("**Percentage by chemical class:**\n")
for (i in seq_len(nrow(compounds_by_class))) {
  cat("- ", as.character(compounds_by_class$new_class[i]), ": ", round(compounds_by_class$pct[i], 1), "%\n", sep = "")
}
voc_per_plant <- df |> group_by(plant) |> summarise(n_voc = n_distinct(compound), .groups = "drop")
max_voc <- max(voc_per_plant$n_voc)
plant_max_voc <- voc_per_plant |> filter(n_voc == max_voc) |> dplyr::pull(plant)
cat("\n**Plant species with the highest diversity of FVOC:** ", paste(plant_max_voc, collapse = ", "), " (", max_voc, " compounds)\n", sep = "")
comp_per_sample <- df |> group_by(unique_id, new_cat) |> summarise(n_comp = n_distinct(compound), .groups = "drop")
comp_summary <- comp_per_sample |> group_by(new_cat) |> summarise(median_comp = median(n_comp), min_comp = min(n_comp), max_comp = max(n_comp), .groups = "drop")
cat("\n**Median compound composition (compounds per sample)** and range:\n")
for (i in seq_len(nrow(comp_summary))) {
  cat("- ", as.character(comp_summary$new_cat[i]), ": median = ", round(comp_summary$median_comp[i], 1), ", range = ", comp_summary$min_comp[i], "–", comp_summary$max_comp[i], "\n", sep = "")
}
```

### ESM1
```{r}
#| label: ESM1
#| echo: false
#| fig.width: 9
#| fig.height: 6
#| fig-cap: "ESM 1. Total number of unique floral volatile organic compounds identified per plant family sampled in Aotearoa-New Zealand."

# Compounds per plant family (bars ordered by count)
df |> group_by(family) |> summarise(count_comp = n_distinct(compound), .groups = "drop") |>
  mutate(family = fct_reorder(family, count_comp)) |>
  ggplot(aes(x = family, y = count_comp)) + geom_col() + geom_hline(yintercept = 100, linetype = "dashed") + coord_flip() +
  ggtitle("Number of compounds per plant family") + theme_classic() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
ggsave(here::here("Plots", "ESM1.png"), width = 9, height = 6, dpi = 600)
```

### ESM2
```{r}
#| label: ESM2
#| echo: false
#| fig.width: 9
#| fig.height: 6
#| fig-cap: "ESM 2. Relative proportions of each chemical class per plant family."

# Chemical class proportions per plant family (Benzenoid grey, Terpenoids red, Fatty acid derivatives blue, NA black)
ggplot(df, aes(x = family, fill = new_class)) + geom_bar(position = "fill") +
  scale_fill_manual(name = "Compound Class", values = c("Benzenoid" = "#A8A6A7FF", "Fatty acid derivatives" = "#006A8EFF", "Terpenoids" = "#B1283AFF", "NA" = "black")) +
  ggtitle("Chemical class per plant family") + coord_flip() + scale_y_reverse() +
  theme_classic() + theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(y = "Relative proportions", fill = "Chemical class")
ggsave(here::here("Plots", "ESM2.png"), width = 9, height = 6, dpi = 600)
```

### ESM3
```{r}
#| label: ESM3
#| echo: false
#| fig.width: 9
#| fig.height: 6
#| fig-cap: "ESM 3. Number of floral volatile organic compounds per chemical sub-class among native, ornamental and crop species from Aotearoa-New Zealand."

# Chemical sub-class proportions by category; NA = black
mycolors <- c(brewer.pal(name = "Spectral", n = 11), brewer.pal(name = "Dark2", n = 3), "black")
df$new_sub_class <- factor(df$new_sub_class, levels = c("Fatty alcohols", "Fatty aldehydes", "Fatty esters", "Fatty unknown", "Hydrocarbons", "Lactones", "Monoterpenes", "Sesquiterpenes", "Irregular terpenes", "Terpenoids unknown", "Phenolic acids", "Phenylethanoids", "Phenylpropanoids", "Benzenoids unknown", "NA"))
ggplot(df, aes(x = new_cat, fill = new_sub_class)) + geom_bar(position = "fill") +
  ggtitle("Chemical sub-class per category") + coord_flip() +
  labs(x = "Origin", y = "Proportions", fill = "Chemical sub-class") +
  theme_classic() + theme(axis.text.x = element_text(angle = 90, hjust = 1), axis.text.y = element_text(face = "bold", size = 10)) +
  scale_fill_manual(values = mycolors)
ggsave(here::here("Plots", "ESM3.png"), width = 9, height = 6, dpi = 600)
```

::: {.callout-note title="ESM3: Count (n) by category and chemical sub-class (click to show/hide)" collapse="true"}
```{r}
#| label: esm3_table

esm3_counts <- df |> count(new_cat, new_sub_class, name = "n") |> rename(Category = new_cat, `Chemical sub-class` = new_sub_class)
esm3_counts
```
:::

### ESM4
```{r}
#| label: ESM4
#| echo: false
#| fig.width: 9
#| fig.height: 6
#| fig-cap: "ESM 4. Non-metric multidimensional scaling (NMDS) representation of the floral volatile organic compounds (FVOCs) composition of each sample for each plant (represented by a point) among the three categories: crops (light yellow), ornamentals (orange) and natives (green) sampled in Aotearoa-New Zealand."

# NMDS on plant × compound matrix (Bray–Curtis); exclude ornamental outliers
df_wide_full <- df |> reshape2::dcast(plant + new_cat + family + order ~ compound, value.var = "abundance", fun.aggregate = sum)
excluded_idx <- c(8, 47, 67, 102)
excluded_plants <- df_wide_full |> slice(excluded_idx) |> dplyr::pull(plant)
df_wide <- df_wide_full |> slice(-excluded_idx)
df_compounds <- df_wide |> dplyr::select(5:ncol(df_wide))
invisible(capture.output(nmds_result <- metaMDS(df_compounds, k = 2, trymax = 50)))
nmds_scores <- vegan::scores(nmds_result, display = "sites") |> as.data.frame() |> mutate(new_cat = df_wide$new_cat)

ggplot(nmds_scores, aes(x = NMDS1, y = NMDS2, color = new_cat)) +
  geom_point(size = 3, alpha = 0.8) + stat_ellipse(linetype = 2, size = 1) +
  scale_color_manual(values = c("#DDCC77", "darkgreen", "orange")) +
  labs(title = "NMDS Ordination of VOC Compounds", subtitle = paste0("Stress = ", round(nmds_result$stress, 3), " | n = ", nrow(nmds_scores)), x = "NMDS1", y = "NMDS2", color = "Category") +
  theme_minimal()
ggsave(here::here("Plots", "ESM4.png"), width = 9, height = 6, dpi = 600)
```

Observations excluded from NMDS: `r paste(excluded_plants, collapse = ", ")`

::: {.callout-note title="NMDS including removed observations (click to show/hide)" collapse="true"}
```{r}
#| label: nmds_all_obs
#| fig.width: 9
#| fig.height: 6

# NMDS on full plant × compound matrix (no exclusions)
df_compounds_full <- df_wide_full |> dplyr::select(5:ncol(df_wide_full))
invisible(capture.output(nmds_full <- metaMDS(df_compounds_full, k = 2, trymax = 50)))
nmds_scores_full <- vegan::scores(nmds_full, display = "sites") |> as.data.frame() |> mutate(new_cat = df_wide_full$new_cat)

ggplot(nmds_scores_full, aes(x = NMDS1, y = NMDS2, color = new_cat)) +
  geom_point(size = 3, alpha = 0.8) + stat_ellipse(linetype = 2, size = 1) +
  scale_color_manual(values = c("#DDCC77", "darkgreen", "orange")) +
  labs(title = "NMDS Ordination (all observations)", subtitle = paste0("Stress = ", round(nmds_full$stress, 3), " | n = ", nrow(nmds_scores_full)), x = "NMDS1", y = "NMDS2", color = "Category") +
  theme_minimal()
```
:::

::: {.callout-note title="ESM4: PERMANOVA (Bray–Curtis, category)" collapse="true"}
```{r}
#| label: stats_esm4

print(adonis2(df_compounds ~ new_cat, data = df_wide, method = "bray", na.action = na.fail))
```
:::

### ESM5
```{r}
#| label: ESM5
#| fig.width: 9
#| fig.height: 4
#| fig-cap: "ESM 5. Boxplots of distances of plant samples to their group centroid based on Bray–Curtis dissimilarities for each plant category. Larger values indicate greater within-group dispersion in floral volatile organic compounds (FVOCs) composition."

# Multivariate dispersion (betadisper); uses df_wide/df_compounds from ESM4
bd <- betadisper(vegdist(df_compounds, method = "bray"), df_wide$new_cat)
boxplot(bd, xlab = "Plant category", ylab = "Distance to centroid (Bray–Curtis)", main = "Multivariate dispersion", col = c("#DDCC77", "darkgreen", "orange"))
invisible(capture.output({
  png(here::here("Plots", "ESM5.png"), width = 9, height = 4, units = "in", res = 600)
  boxplot(bd, xlab = "Plant category", ylab = "Distance to centroid (Bray–Curtis)", main = "Multivariate dispersion", col = c("#DDCC77", "darkgreen", "orange"))
  dev.off()
}))
```

::: {.callout-note title="ESM5: Dispersion (anova/permutest) and ANOVA (richness, Shannon)" collapse="true"}
```{r}
#| label: stats_esm5

anova(bd)
permutest(bd, permutations = 999)
richness <- specnumber(df_compounds)
summary(aov(richness ~ new_cat, data = df_wide))
shannon_div <- vegan::diversity(df_compounds)
summary(aov(shannon_div ~ new_cat, data = df_wide))
```
:::

### ESM6
```{r}
#| label: ESM6
#| fig.width: 14
#| fig.height: 8
#| fig-cap: "ESM 6. Phylogenetic tree for the 113 plants sampled in Aotearoa-New Zealand and colored by categories: crops (light yellow), ornamentals (orange) and natives (green)."

# --- Build phylogeny (V.PhyloMaker2) ---
df$species <- df$plant
df <- df |> mutate(family = as.character(family), family = if_else(genus == "Griselinia", "Griseliniaceae", family), family = factor(family))
df <- df |> mutate(family = as.character(family), family = if_else(genus == "Phormium", "Asphodelaceae", family), family = factor(family))
sp.list <- df |> dplyr::distinct(genus, species, family, new_cat) |> dplyr::select(species, genus, family, new_cat)
data("GBOTB.extended.TPL")
data("nodes.info.1.TPL")
tree <- phylo.maker(sp.list = sp.list, tree = GBOTB.extended.TPL, nodes = nodes.info.1.TPL, scenarios = "S3")
my_tree <- tree$scenario.3
my_tree$tip.label <- gsub(" ", "_", my_tree$tip.label)
sp.list_clean <- sp.list |> dplyr::mutate(label = gsub(" ", "_", as.character(species))) |> dplyr::select(label, new_cat)

# Circular phylogeny: tips coloured by plant category (Crops, Natives, Ornamentals)
ggtree(my_tree, layout = "circular") %<+% sp.list_clean +
  geom_tippoint(aes(color = new_cat), size = 2, na.rm = TRUE) +
  geom_tiplab(aes(color = new_cat), size = 3, offset = 0.5) +
  labs(color = "Plant Category") +
  scale_color_manual(values = c(Crops = "#DDCC77", Natives = "darkgreen", Ornamentals = "orange"), na.value = "grey50") +
  theme_void() + theme(legend.justification = c(1, 1))

# --- Prune tree to species in VOC data; prepare for Pagel's λ and PGLS ---
voc_richness <- df |> group_by(plant) |> summarise(voc_richness = n_distinct(compound), new_cat = dplyr::first(as.character(new_cat)), .groups = "drop")
voc_richness$plant <- gsub(" ", "_", voc_richness$plant)
common <- dplyr::intersect(my_tree$tip.label, voc_richness$plant)
tree2 <- ape::drop.tip(my_tree, dplyr::setdiff(my_tree$tip.label, common))
voc_richness <- voc_richness |> filter(plant %in% common) |> slice(match(tree2$tip.label, plant))
voc_richness_df <- as.data.frame(voc_richness)
rownames(voc_richness_df) <- NULL
tree2$node.label <- NULL
voc_richness_df$plant <- as.character(voc_richness_df$plant)
ggsave(here::here("Plots", "ESM6.png"), width = 14, height = 8, dpi = 600)
```

::: {.callout-note title="ESM6: Pagel's λ (phylogenetic signal in VOC richness)" collapse="true"}
```{r}
#| label: ESM6_pagel

lambda <- phytools::phylosig(tree2, voc_richness$voc_richness, method = "lambda", test = TRUE)
print(lambda)
```
:::

::: {.callout-note title="ESM6: PGLS (VOC richness ~ intercept) & (VOC richness ~ category)" collapse="true"}
```{r}
#| label: ESM6_pgls

comp_data <- caper::comparative.data(phy = tree2, data = voc_richness_df, names.col = "plant", vcv = TRUE)
pgls_model_intercept <- caper::pgls(voc_richness ~ 1, data = comp_data, lambda = "ML")
pgls_model_category <- caper::pgls(voc_richness ~ new_cat, data = comp_data, lambda = "ML")
summary(pgls_model_intercept)
summary(pgls_model_category)
```
:::

::: {.callout-note title="ESM6: Phylogenetic distance vs VOC distance (Mantel test)" collapse="true"}
```{r}
#| label: ESM6_mantel

# Species × compound matrix (abundance) for species in tree2
voc_wide <- df |> mutate(plant = gsub(" ", "_", plant)) |> group_by(plant, compound) |> summarise(abundance = sum(abundance), .groups = "drop")
voc_matrix <- voc_wide |> reshape2::dcast(plant ~ compound, value.var = "abundance", fill = 0)
rownames(voc_matrix) <- voc_matrix$plant
voc_matrix <- as.matrix(voc_matrix[, -1])
common_sp <- intersect(tree2$tip.label, rownames(voc_matrix))
voc_matrix <- voc_matrix[common_sp, , drop = FALSE]
tree2_mantel <- ape::drop.tip(tree2, setdiff(tree2$tip.label, common_sp))
voc_dist <- vegdist(voc_matrix, method = "bray")
phylo_dist <- cophenetic(tree2_mantel)
mantel(voc_dist, phylo_dist, permutations = 999)
```
:::

### ESM7
```{r}
#| label: regional_crops_bar
#| fig.width: 9
#| fig.height: 4.5
#| fig-cap: "ESM 7. Top 10 most frequent compounds for exotic crops (arable and fruit) grown in Aotearoa-New Zealand."

# Top 10 VOCs per crop type (Arable vs Fruit); uses df_crops from Figure 6
df_ranked_crops <- df_crops |> group_by(type) |> count(compound, sort = TRUE) |>
  slice_max(n, n = 10, with_ties = FALSE) |> ungroup()

ggplot(df_ranked_crops, aes(x = tidytext::reorder_within(compound, n, type), y = n, fill = type)) +
  geom_bar(stat = "identity", alpha = 0.6) + coord_flip() + facet_wrap(~type, scales = "free", ncol = 2) +
  tidytext::scale_x_reordered() + scale_fill_manual(values = c("darkblue", "darkred")) +
  labs(x = "VOC Compound", y = "Count") + theme_classic() + theme(legend.position = "none")
ggsave(here::here("Plots", "ESM7.png"), width = 9, height = 4.5, dpi = 600)
```

# References

R Core Team (2022). R: A language and environment for statistical computing. R Foundation for Statistical Computing, Vienna, Austria. <https://www.R-project.org/>

::: {.callout-tip title="Session Information" collapse="true"}
```{r}
#| label: session-info

utils::sessionInfo()
```
::: 
