---
title: "NZ Floral Volatilome Analysis"
author: "Tom Moore & Flore Mas"
date: last-modified
output-file: "2_voc_analysis.html"
format:
  html:
    toc: true
    toc-depth: 5
    toc-title: "Contents"
    theme: united
    highlight-style: tango
    fig-width: 9
    fig-height: 6
    code-tools: true
    code-fold: false
    code-overflow: scroll
    df-print: paged
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| results: false

# --- Packages: install if missing, then load ---
if (!requireNamespace("pak", quietly = TRUE)) install.packages("pak")
packages <- c(
  "tidyverse", "readxl", "here", "vegan", "janitor", "tidytext", "ggvenn",
  "reshape2", "remotes", "devtools", "ape", "ggtree", "caper", "phytools",
  "patchwork", "RColorBrewer", "geomorph"
)
missing_pkgs <- packages[!sapply(packages, requireNamespace, quietly = TRUE)]
if (length(missing_pkgs) > 0) {
  pak::pkg_install(missing_pkgs, ask = FALSE)
  cat("Installed:", paste(missing_pkgs, collapse = ", "), "\n")
}
invisible(sapply(packages, library, character.only = TRUE))

# GitHub packages (ggvegan, V.PhyloMaker2)
if (!requireNamespace("ggvegan", quietly = TRUE)) remotes::install_github("gavinsimpson/ggvegan")
if (!requireNamespace("V.PhyloMaker2", quietly = TRUE)) devtools::install_github("jinyizju/V.PhyloMaker2")
library("ggvegan")
library("V.PhyloMaker2")
```

# Overview

This report analyses floral volatile organic compounds (VOCs) from New Zealand plants. We use a cleaned dataset of compounds by sample, with plant species classified as **Natives**, **Crops**, or **Ornamentals**. The analysis covers: (1) chemical class distribution across categories, (2) diversity metrics (richness, Shannon/Hill, evenness) and their comparison by category, (3) overlap of compounds between categories (Venn), (4) seasonal phenology of VOC composition, and (5) supplementary ordination (NMDS), dispersion, and phylogeny. Statistical tests include chi-square (class vs category), ANOVA and post-hoc Wilcoxon (diversity), and Kruskal–Wallis (phenology); test outputs are in collapsible sections so you can show or hide them.

Although AI was used to format the analysis and this Quarto document, all decisions about what types of analysis to perform and what types of plots to use were made by the authors.

::: {.callout-note title="Dataframe (click to show/hide)" collapse="true"}
## Dataframe

```{r}
#| label: data_cleaning
#| echo: false

# --- Load and clean raw VOC data ---
# Load raw VOC data and coerce to factors
df <- read_excel(here::here("Analysis/Data/voc_ldata.xlsx"), sheet = "LDATA_ALL_tous3")
df <- as.data.frame(unclass(df), stringsAsFactors = TRUE)

# Clean names and standardise factor labels
df <- df |> clean_names() |> as.data.frame(stringsAsFactors = TRUE) |>
  mutate(across(where(is.factor), ~ as.factor(str_to_sentence(as.character(.)))))

# Numeric columns
cols_numeric <- c("sum_area", "area", "tot_voc", "n_voc")
df[, cols_numeric] <- lapply(df[, cols_numeric], as.numeric)

# Sample-level ID and category (Natives / Crops / Ornamentals)
df <- df |> mutate(
  unique_id = as.factor(paste(as.character(sample), substr(as.character(location), 1, 3), sep = "-")),
  new_cat = case_when(
    native_exotic == "Native" ~ "Natives",
    native_exotic == "Exotic" & crop_wild == "Crop" ~ "Crops",
    native_exotic == "Exotic" & crop_wild == "Wild" ~ "Ornamentals",
    TRUE ~ NA_character_
  ),
  new_cat = as.factor(new_cat)
)

# Chemical class labels: rename and set order
df <- df |> mutate(
  new_class = fct_recode(new_class, "Fatty acid derivatives" = "Fatty_acid_derivatives"),
  new_sub_class = fct_recode(new_sub_class,
    "Benzenoids unknown" = "Benzenoids_unknown", "Fatty unknown" = "Fatty_unknown",
    "Fatty alcohols" = "Fatty_alcohols", "Fatty aldehydes" = "Fatty_aldehydes",
    "Fatty esters" = "Fatty_esters", "Phenolic acids" = "Phenolic_acids",
    "Irregular terpenes" = "Irregular_terpenes", "Terpenoids unknown" = "Terpenoids-unknown")
)
df$new_class <- factor(df$new_class, levels = c("Fatty acid derivatives", "Terpenoids", "Benzenoid", "NA"))
df$new_sub_class <- factor(df$new_sub_class,
  levels = c("Hydrocarbons", "Fatty esters", "Fatty aldehydes", "Fatty alcohols", "Fatty unknown", "Lactones",
    "Monoterpenes", "Sesquiterpenes", "Irregular terpenes", "Terpenoids unknown", "Benzenoids unknown",
    "Phenylpropanoids", "Phenolic acids", "Phenylethanoids", "NA"))

# Keep required columns and compute relative abundance per sample
df <- df |> dplyr::select(compound, unique_id, plant, type, new_cat, genus, family, order,
  location, region, new_class, new_sub_class, sum_area, perc_area, prop_area, smiles, inchikeys)
df <- df |> group_by(compound, unique_id) |> mutate(abundance = ifelse(is.na(sum_area), perc_area, sum_area)) |>
  ungroup() |> group_by(unique_id) |> mutate(sum_abund = sum(abundance)) |> ungroup() |>
  group_by(compound, unique_id) |> mutate(rel_abund = abundance / sum_abund) |> ungroup()

# Reclassify Trifolium/Lavandula as Ornamentals
df <- df |> mutate(
  new_cat = case_when(
    plant %in% c("Trifolium incarnatum", "Trifolium pratense", "Trifolium repens",
                 "Lavandula angustifolia", "Lavandula dentata") ~ "Ornamentals",
    TRUE ~ as.character(new_cat)
  ),
  new_cat = factor(new_cat)
)
```

**Dataframe structure** (`str(df)`):

```{r}
#| label: str_df
#| echo: false

# Show structure of main dataframe
str(df)
```
:::

::: {.callout-note title="Descriptive statistics (click to show/hide)" collapse="true"}
## Descriptive statistics

Summaries by sample, plant, category, and chemical class (shown below).

```{r}
#| label: descriptive_statistics
#| echo: false

# --- Descriptive summaries ---
# By sample (unique_id): compound count and total VOC
summary_unique_id <- df |> group_by(unique_id) |>
  summarise(unique_id_compound = n_distinct(compound), total_voc = sum(abundance))

# By plant: compound count, replicates, mean total VOC
summary_plant <- df |> group_by(unique_id) |>
  mutate(unique_id_compound = n_distinct(compound), total_voc = sum(abundance)) |>
  ungroup() |> group_by(plant) |>
  summarise(type = first(new_cat), count_compound = n_distinct(compound),
    n_replicates = n_distinct(unique_id), mean_total_voc = mean(total_voc), .groups = "drop")

# By category (new_cat): plants, compound stats, total VOC range
summary_categories <- df |> group_by(unique_id) |>
  mutate(unique_id_compound = n_distinct(compound), total_voc = sum(abundance)) |>
  ungroup() |> group_by(plant) |>
  mutate(count_compound = n_distinct(compound), nb_replicates = n_distinct(unique_id), mean_total_voc = mean(total_voc)) |>
  ungroup() |> group_by(new_cat) |>
  summarise(nb_plant = n_distinct(plant), mean_comp = mean(count_compound), median_voc = median(count_compound),
    min_comp = min(count_compound), max_comp = max(count_compound),
    mean_total_voc = mean(mean_total_voc), min_total_voc = min(mean_total_voc), max_total_voc = max(mean_total_voc))

# By chemical class: count and percent
summary_chem_class <- df |> group_by(new_class) |> summarise(count = n()) |> mutate(percent = 100 * count / sum(count))

# Display tables in report
head(summary_unique_id, 10)
summary_plant
summary_categories
summary_chem_class
```
:::

# Figures & hypothesis tests

## Figure 1
```{r}
#| label: fvoc_frequency
#| fig.width: 9
#| fig.height: 6

# Top 10 compounds by chemical class (Benzenoid, Fatty acid derivatives, Terpenoids)
df_ranked_class <- df |> group_by(new_class) |> count(compound, sort = TRUE) |>
  slice_max(n, n = 10, with_ties = FALSE) |> ungroup() |>
  filter(new_class != "NA") |> droplevels() |>
  mutate(new_class = fct_relevel(new_class, "Benzenoid", "Fatty acid derivatives", "Terpenoids"))

ggplot(df_ranked_class, aes(x = reorder(compound, n), y = n, fill = new_class)) +
  geom_bar(stat = "identity", alpha = 0.6) + coord_flip() +
  scale_fill_manual(name = "Compound Class", values = c("#A8A6A7FF", "#006A8EFF", "#B1283AFF")) +
  labs(x = "Floral Volatile Organic Compounds", y = "Count") +
  theme_classic() + theme(legend.position = "right")
ggsave(here::here("Plots", "Figure1.png"), width = 9, height = 6, dpi = 600)
```

## Figure 2
```{r}
#| label: chemical_class_distribution
#| fig.width: 9
#| fig.height: 3

# Chemical class proportions by category (donut-style pie charts)
df_pie <- df |> filter(new_class != "NA") |> droplevels() |>
  group_by(new_cat, new_class) |> count() |> group_by(new_cat) |>
  mutate(percent = (n / sum(n)) * 100) |> ungroup()
df_pie$new_class <- factor(df_pie$new_class, levels = c("Benzenoid", "Fatty acid derivatives", "Terpenoids"))

ggplot(df_pie, aes(x = "", y = percent, fill = new_class)) +
  geom_bar(stat = "identity", width = 1, alpha = 0.8, aes(color = new_cat), linewidth = 1.5) +
  coord_polar("y", start = 0, direction = -1) + facet_wrap(~new_cat, ncol = 3) +
  scale_fill_manual(name = "Compound Class", values = c("#A8A6A7FF", "#006A8EFF", "#B1283AFF")) +
  scale_color_manual(values = c("Crops" = "#DDCC77", "Natives" = "darkgreen", "Ornamentals" = "orange"), guide = "none") +
  theme_void() + theme(legend.position = "bottom", strip.text = element_text(size = 12, face = "bold"))
ggsave(here::here("Plots", "Figure2.png"), width = 9, height = 3, dpi = 600)
```

::: {.callout-note title="Chi-square test (chemical class vs category)" collapse="true"}
```{r}
#| label: chisq_fig2

contingency_table <- df |> filter(new_class != "NA") |> count(new_cat, new_class) |>
  pivot_wider(names_from = new_class, values_from = n, values_fill = 0) |> column_to_rownames("new_cat")
print(chisq.test(contingency_table))
```
:::

## Figure 3
```{r}
#| label: ven_diagram
#| fig.width: 9
#| fig.height: 6

# Venn: unique compounds per category (Crops, Natives, Ornamentals)
venn_data <- df |> group_by(new_cat) |> summarise(compounds = list(unique(compound))) |> deframe()
ggvenn(venn_data, fill_color = c("#DDCC77", "darkgreen", "orange"), fill_alpha = 0.5, stroke_size = 0.5, set_name_size = 5, text_size = 3)
ggsave(here::here("Plots", "Figure3.png"), width = 9, height = 6, dpi = 600)
```

## Figure 4
```{r}
#| label: top_10
#| fig.width: 9
#| fig.height: 3.5

# Top 10 VOCs per category
df_ranked <- df |> group_by(new_cat) |> count(compound, sort = TRUE) |>
  slice_max(n, n = 10, with_ties = FALSE) |> ungroup()

ggplot(df_ranked, aes(x = reorder_within(compound, n, new_cat), y = n, fill = new_cat)) +
  geom_bar(stat = "identity", alpha = 0.6) + coord_flip() + facet_wrap(~new_cat, ncol = 3, scales = "free") +
  scale_x_reordered() + scale_fill_manual(values = c("#DDCC77", "darkgreen", "orange")) +
  labs(x = "Top 10 Volatile Organic Compounds", y = "Count") +
  theme_classic() + theme(legend.position = "none", strip.text = element_text(size = 12, face = "bold"))
ggsave(here::here("Plots", "Figure4.png"), width = 9, height = 3.5, dpi = 600)
```

## Figure 5
```{r}
#| label: hills_diversity
#| fig.width: 9
#| fig.height: 3.5

# Wide format: one row per sample, columns = compounds (abundance)
df_wide <- df |> reshape2::dcast(unique_id + new_cat + location + region + family + order + plant ~ compound, value.var = "abundance", fun.aggregate = sum)
df_compounds <- df_wide |> dplyr::select(where(is.numeric))

# Diversity: richness (q=0), Shannon/Hill (q=1), evenness (Pielou J)
richness <- specnumber(df_compounds)
shannon_div <- vegan::diversity(df_compounds, index = "shannon")
hill_q1 <- exp(shannon_div)
evenness <- shannon_div / log(richness)
diversity_df <- data.frame(
  unique_id = df_wide$unique_id, new_cat = df_wide$new_cat, family = df_wide$family,
  order = df_wide$order, plant = df_wide$plant, richness = richness, hill_q1 = hill_q1, evenness = evenness
)

# Boxplots (Richness, Evenness, Hill) by category
pal <- c("#DDCC77", "darkgreen", "orange")
p_richness <- ggplot(diversity_df, aes(x = new_cat, y = richness, fill = new_cat, color = new_cat)) +
  geom_jitter(alpha = 0.1, width = 0.3) + geom_boxplot(width = 0.3, alpha = 0.5, outlier.shape = NA) +
  scale_y_continuous(breaks = seq(0, max(diversity_df$richness), by = 10)) +
  scale_color_manual(values = pal) + scale_fill_manual(values = pal) +
  labs(y = "Richness", x = "") + theme_bw() +
  theme(legend.position = "none", axis.text = element_text(size = 10), axis.title = element_text(size = 11, face = "bold"))
p_evenness <- ggplot(diversity_df, aes(x = new_cat, y = evenness, fill = new_cat, color = new_cat)) +
  geom_jitter(alpha = 0.1, width = 0.3) + geom_boxplot(width = 0.3, alpha = 0.5, outlier.shape = NA) +
  scale_y_continuous(limits = c(0, 1.01), breaks = seq(0, 1, 0.2)) +
  scale_color_manual(values = pal) + scale_fill_manual(values = pal) +
  labs(y = "Evenness", x = "") + theme_bw() +
  theme(legend.position = "none", axis.text = element_text(size = 10), axis.title = element_text(size = 11, face = "bold"))
p_hill <- ggplot(diversity_df, aes(x = new_cat, y = hill_q1, fill = new_cat, color = new_cat)) +
  geom_jitter(alpha = 0.1, width = 0.3) + geom_boxplot(width = 0.3, alpha = 0.5, outlier.shape = NA) +
  scale_color_manual(values = pal) + scale_fill_manual(values = pal) +
  labs(y = "Hill Diversity", x = "") + theme_bw() +
  theme(legend.position = "none", axis.text = element_text(size = 10), axis.title = element_text(size = 11, face = "bold"))
p_richness + p_evenness + p_hill + plot_layout(ncol = 3) + plot_annotation(tag_levels = "A")
ggsave(here::here("Plots", "Figure5.png"), width = 9, height = 3.5, dpi = 600)
```

::: {.callout-note title="Figure 5: ANOVA and post-hoc Wilcoxon (diversity by category)" collapse="true"}
```{r}
#| label: stats_fig5

summary(aov(richness ~ new_cat, data = diversity_df))
summary(aov(hill_q1 ~ new_cat, data = diversity_df))
summary(aov(evenness ~ new_cat, data = diversity_df))
pairwise.wilcox.test(diversity_df$richness, diversity_df$new_cat, p.adjust.method = "BH")
pairwise.wilcox.test(diversity_df$hill_q1, diversity_df$new_cat, p.adjust.method = "BH")
pairwise.wilcox.test(diversity_df$evenness, diversity_df$new_cat, p.adjust.method = "BH")
```
:::

::: {.callout-note title="Figure 5: Disparity analysis (mean distance by category)" collapse="true"}
```{r}
#| label: disparity_fig5

# Disparity analysis: mean within- and between-group distances (Euclidean on compound abundances)
df_compounds <- df_wide |> dplyr::select(where(is.numeric))
dist_matrix <- dist(df_compounds, method = "euclidean")
disparity <- vegan::meandist(dist_matrix, df_wide$new_cat)
disparity
plot(disparity, kind = "histogram")
```
:::

## Figure 6
```{r}
#| label: regional_crops
#| fig.width: 9
#| fig.height: 4.5

# Crops only: chemical class proportions by region
df_crops <- df |> filter(type %in% c("Arable", "Fruit"), !is.na(new_class)) |> droplevels()
df_pie <- df_crops |> group_by(region, new_class) |> count() |> group_by(region) |>
  mutate(percent = (n / sum(n)) * 100) |> ungroup()
df_pie <- df_pie |> mutate(
  region = fct_recode(region,
    "Tropical fruit" = "Central north island",
    "Pipfruit & Stonefruit" = "Lower north island",
    "Arable crops" = "Upper south island",
    "Summerfruit" = "Lower south island",
    "Natives" = "Aotearoa"),
  new_class = fct_relevel(new_class, "Benzenoid", "Fatty acid derivatives", "Terpenoids")
)

ggplot(df_pie, aes(x = "", y = percent, fill = new_class)) +
  geom_bar(stat = "identity", alpha = 0.8) + coord_polar("y", start = 0, direction = -1) +
  facet_wrap(~region) +
  scale_fill_manual(name = "Compound Class", values = c("#A8A6A7FF", "#006A8EFF", "#B1283AFF")) +
  theme_void() + theme(legend.position = "bottom", legend.title = element_blank(), strip.text = element_text(size = 10))
ggsave(here::here("Plots", "Figure6.png"), width = 9, height = 4.5, dpi = 600)
```

## Figure 7
```{r}
#| label: phenology
#| fig.width: 9
#| fig.height: 6.5

# Phenology: flowering season per plant; merge with VOC data
pheno <- read_excel(here::here("Analysis/Data/flower_phenology.xlsx"), sheet = "Sheet1")
pheno <- as.data.frame(unclass(pheno), stringsAsFactors = TRUE) |> clean_names() |>
  mutate(across(where(is.factor), ~ as.factor(str_to_sentence(as.character(.)))))
df_pheno <- df |> left_join(pheno, by = join_by(plant), relationship = "many-to-many") |> droplevels()
df_pheno <- df_pheno |> mutate(season = fct_recode(season, "Autumn" = "Fall")) |> droplevels()
df_pheno$season <- factor(df_pheno$season, levels = c("Spring", "Summer", "Autumn", "All-year"))

# Filter: exclude All-year and NA chemical class; mean rel_abund by category × class × season
df_pheno_filtered <- df_pheno |> filter(season != "All-year" & new_class != "NA") |> droplevels() |>
  mutate(new_class = fct_relevel(new_class, "Benzenoid", "Fatty acid derivatives", "Terpenoids"))
mean_df <- df_pheno_filtered |> group_by(new_cat, new_class, season) |>
  summarise(mean_abund = mean(rel_abund, na.rm = TRUE), .groups = "drop")
count_df <- df_pheno_filtered |> group_by(new_cat, season) |> summarise(nb_plant = n_distinct(plant), .groups = "drop")

# Panel A: mean relative abundance by season and chemical class (by category)
p1 <- ggplot(df_pheno_filtered, aes(x = season, y = rel_abund, colour = new_class, group = new_class)) +
  geom_smooth(data = mean_df, aes(x = season, y = mean_abund, colour = new_class, group = new_class), method = "loess", span = 1, se = FALSE) +
  geom_point(data = mean_df, aes(x = season, y = mean_abund, colour = new_class), size = 2.5) +
  facet_grid(new_cat ~ .) +
  scale_colour_manual(name = "Chemical Class", values = c("Benzenoid" = "#A8A6A7FF", "Fatty acid derivatives" = "#006A8EFF", "Terpenoids" = "#B1283AFF")) +
  labs(title = "A) Seasonal patterns of chemical classes", y = "Mean relative abundance", x = "Season") +
  theme_classic() + theme(legend.position = "right", strip.text = element_text(size = 12, face = "bold"))
# Panel B: number of plants flowering per season
p2 <- ggplot(count_df, aes(x = season, y = nb_plant, fill = new_cat)) +
  geom_bar(stat = "identity", position = "dodge", width = 0.6, alpha = 0.8) +
  scale_fill_manual(name = "Category", values = c("#DDCC77", "darkgreen", "orange")) +
  labs(title = "B) Number of flowering plants per season", x = "Season", y = "Number of plants") +
  theme_classic() + theme(legend.position = "right", strip.text = element_text(size = 12, face = "bold"))
(p1 / p2) + plot_layout(heights = c(3, 1))
ggsave(here::here("Plots", "Figure7.png"), width = 9, height = 6.5, dpi = 600)
```

::: {.callout-note title="Figure 7: Kruskal–Wallis and post-hoc Wilcoxon (phenology by category × class)" collapse="true"}
```{r}
#| label: stats_phenology

for (cat in levels(df_pheno_filtered$new_cat)) {
  cat_data <- df_pheno_filtered |> filter(new_cat == cat)
  for (class in levels(cat_data$new_class)) {
    class_data <- cat_data |> filter(new_class == class)
    if (nrow(class_data) > 0 & length(unique(class_data$season)) > 1) {
      kw_test <- kruskal.test(rel_abund ~ season, data = class_data)
      cat(cat, "-", class, ": Kruskal-Wallis p =", format.pval(kw_test$p.value, digits = 3), "\n")
      if (kw_test$p.value < 0.05) print(pairwise.wilcox.test(class_data$rel_abund, class_data$season, p.adjust.method = "BH"))
    }
  }
}
```
:::

# Supplementary material

### ESM1
```{r}
#| label: ESM1
#| echo: false
#| fig.width: 9
#| fig.height: 6

# Compounds per plant family (bars ordered by count)
df |> group_by(family) |> summarise(count_comp = n_distinct(compound), .groups = "drop") |>
  mutate(family = fct_reorder(family, count_comp)) |>
  ggplot(aes(x = family, y = count_comp)) + geom_col() + coord_flip() +
  ggtitle("Number of compounds per plant family") + theme_classic() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
ggsave(here::here("Plots", "ESM1.png"), width = 9, height = 6, dpi = 600)
```

### ESM2
```{r}
#| label: ESM2
#| echo: false
#| fig.width: 9
#| fig.height: 6

# Chemical class proportions per plant family (Benzenoid grey, Terpenoids red, Fatty acid derivatives blue, NA black)
ggplot(df, aes(x = family, fill = new_class)) + geom_bar(position = "fill") +
  scale_fill_manual(name = "Compound Class", values = c("Benzenoid" = "#A8A6A7FF", "Fatty acid derivatives" = "#006A8EFF", "Terpenoids" = "#B1283AFF", "NA" = "black")) +
  ggtitle("Chemical class per plant family") + coord_flip() + scale_y_reverse() +
  theme_classic() + theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(y = "Relative proportions", fill = "Chemical class")
ggsave(here::here("Plots", "ESM2.png"), width = 9, height = 6, dpi = 600)
```

### ESM3
```{r}
#| label: ESM3
#| echo: false
#| fig.width: 9
#| fig.height: 6

# Chemical sub-class proportions by category; NA = black
mycolors <- c(brewer.pal(name = "Spectral", n = 11), brewer.pal(name = "Dark2", n = 3), "black")
df$new_sub_class <- factor(df$new_sub_class, levels = c("Fatty alcohols", "Fatty aldehydes", "Fatty esters", "Fatty unknown", "Hydrocarbons", "Lactones", "Monoterpenes", "Sesquiterpenes", "Irregular terpenes", "Terpenoids unknown", "Phenolic acids", "Phenylethanoids", "Phenylpropanoids", "Benzenoids unknown", "NA"))
ggplot(df, aes(x = new_cat, fill = new_sub_class)) + geom_bar(position = "fill") +
  ggtitle("Chemical sub-class per category") + coord_flip() +
  labs(x = "Origin", y = "Proportions", fill = "Chemical sub-class") +
  theme_classic() + theme(axis.text.x = element_text(angle = 90, hjust = 1), axis.text.y = element_text(face = "bold", size = 10)) +
  scale_fill_manual(values = mycolors)
ggsave(here::here("Plots", "ESM3.png"), width = 9, height = 6, dpi = 600)
```

### ESM4
```{r}
#| label: ESM4
#| echo: false
#| fig.width: 9
#| fig.height: 6

# NMDS on plant × compound matrix (Bray–Curtis); exclude ornamental outliers
df_wide <- df |> reshape2::dcast(plant + new_cat + family + order ~ compound, value.var = "abundance", fun.aggregate = sum) |> slice(-c(8, 47, 67, 102))
df_compounds <- df_wide |> dplyr::select(5:ncol(df_wide))
invisible(capture.output(nmds_result <- metaMDS(df_compounds, k = 2, trymax = 50)))
nmds_scores <- vegan::scores(nmds_result, display = "sites") |> as.data.frame() |> mutate(new_cat = df_wide$new_cat)

ggplot(nmds_scores, aes(x = NMDS1, y = NMDS2, color = new_cat)) +
  geom_point(size = 3, alpha = 0.8) + stat_ellipse(linetype = 2, size = 1) +
  scale_color_manual(values = c(Crops = "#DDCC77", Natives = "darkgreen", Ornamentals = "orange")) +
  labs(title = "NMDS Ordination of VOC Compounds", subtitle = paste0("Stress = ", round(nmds_result$stress, 3), " | n = ", nrow(nmds_scores)), x = "NMDS1", y = "NMDS2", color = "Category") +
  theme_minimal()
ggsave(here::here("Plots", "ESM4.png"), width = 9, height = 6, dpi = 600)
```

::: {.callout-note title="ESM4: PERMANOVA (Bray–Curtis, category)" collapse="true"}
```{r}
#| label: stats_esm4

print(adonis2(df_compounds ~ new_cat, data = df_wide, method = "bray", na.action = na.fail))
```
:::

### ESM5
```{r}
#| label: ESM5
#| fig.width: 9
#| fig.height: 4

# Multivariate dispersion (betadisper); uses df_wide/df_compounds from ESM4
bd <- betadisper(vegdist(df_compounds, method = "bray"), df_wide$new_cat)
cat_colours <- c(Crops = "#DDCC77", Natives = "darkgreen", Ornamentals = "orange")
boxplot(bd, xlab = "Plant category", ylab = "Distance to centroid (Bray–Curtis)", main = "Multivariate dispersion", col = cat_colours[levels(df_wide$new_cat)])
png(here::here("Plots", "ESM5.png"), width = 9, height = 4, units = "in", res = 600)
boxplot(bd, xlab = "Plant category", ylab = "Distance to centroid (Bray–Curtis)", main = "Multivariate dispersion", col = cat_colours[levels(df_wide$new_cat)])
dev.off()
```

::: {.callout-note title="ESM5: Dispersion (anova/permutest) and ANOVA (richness, Shannon)" collapse="true"}
```{r}
#| label: stats_esm5

anova(bd)
permutest(bd, permutations = 999)
richness <- specnumber(df_compounds)
summary(aov(richness ~ new_cat, data = df_wide))
shannon_div <- vegan::diversity(df_compounds)
summary(aov(shannon_div ~ new_cat, data = df_wide))
```
:::

### ESM6
```{r}
#| label: ESM6
#| fig.width: 14
#| fig.height: 8

# --- Build phylogeny (V.PhyloMaker2) ---
df$species <- df$plant
df <- df |> mutate(family = as.character(family), family = if_else(genus == "Griselinia", "Griseliniaceae", family), family = factor(family))
df <- df |> mutate(family = as.character(family), family = if_else(genus == "Phormium", "Asphodelaceae", family), family = factor(family))
sp.list <- df |> dplyr::distinct(genus, species, family, new_cat) |> dplyr::select(species, genus, family, new_cat)
data("GBOTB.extended.TPL")
data("nodes.info.1.TPL")
tree <- phylo.maker(sp.list = sp.list, tree = GBOTB.extended.TPL, nodes = nodes.info.1.TPL, scenarios = "S3")
my_tree <- tree$scenario.3
my_tree$tip.label <- gsub(" ", "_", my_tree$tip.label)
sp.list_clean <- sp.list |> dplyr::mutate(label = gsub(" ", "_", as.character(species))) |> dplyr::select(label, new_cat)

# Circular phylogeny: tips coloured by plant category (Crops, Natives, Ornamentals)
ggtree(my_tree, layout = "circular") %<+% sp.list_clean +
  geom_tippoint(aes(color = new_cat), size = 2, na.rm = TRUE) +
  geom_tiplab(aes(color = new_cat), size = 3, offset = 0.5) +
  labs(color = "Plant Category") +
  scale_color_manual(values = c(Crops = "#DDCC77", Natives = "darkgreen", Ornamentals = "orange"), na.value = "grey50") +
  theme_void() + theme(legend.justification = c(1, 1))

# --- Prune tree to species in VOC data; prepare for Pagel's λ and PGLS ---
voc_richness <- df |> group_by(plant) |> summarise(voc_richness = n_distinct(compound), new_cat = dplyr::first(as.character(new_cat)), .groups = "drop")
voc_richness$plant <- gsub(" ", "_", voc_richness$plant)
common <- dplyr::intersect(my_tree$tip.label, voc_richness$plant)
tree2 <- ape::drop.tip(my_tree, dplyr::setdiff(my_tree$tip.label, common))
voc_richness <- voc_richness |> filter(plant %in% common) |> slice(match(tree2$tip.label, plant))
voc_richness_df <- as.data.frame(voc_richness)
rownames(voc_richness_df) <- NULL
tree2$node.label <- NULL
voc_richness_df$plant <- as.character(voc_richness_df$plant)
ggsave(here::here("Plots", "ESM6.png"), width = 14, height = 8, dpi = 600)
```

::: {.callout-note title="ESM6: Pagel's λ (phylogenetic signal in VOC richness)" collapse="true"}
```{r}
#| label: ESM6_pagel

lambda <- phytools::phylosig(tree2, voc_richness$voc_richness, method = "lambda", test = TRUE)
print(lambda)
```
:::

::: {.callout-note title="ESM6: PGLS (VOC richness ~ 1)" collapse="true"}
```{r}
#| label: ESM6_pgls

comp_data <- caper::comparative.data(phy = tree2, data = voc_richness_df, names.col = "plant", vcv = TRUE)
pgls_model <- caper::pgls(voc_richness ~ 1, data = comp_data, lambda = "ML")
summary(pgls_model)
```
:::

### ESM7
```{r}
#| label: regional_crops_bar
#| fig.width: 9
#| fig.height: 4.5

# Top 10 VOCs per crop type (Arable vs Fruit); uses df_crops from Figure 6
df_ranked_crops <- df_crops |> group_by(type) |> count(compound, sort = TRUE) |>
  slice_max(n, n = 10, with_ties = FALSE) |> ungroup()

ggplot(df_ranked_crops, aes(x = tidytext::reorder_within(compound, n, type), y = n, fill = type)) +
  geom_bar(stat = "identity", alpha = 0.6) + coord_flip() + facet_wrap(~type, scales = "free", ncol = 2) +
  tidytext::scale_x_reordered() + scale_fill_manual(values = c("darkblue", "darkred")) +
  labs(x = "VOC Compound", y = "Count") + theme_classic() + theme(legend.position = "none")
ggsave(here::here("Plots", "ESM7.png"), width = 9, height = 4.5, dpi = 600)
```

# References

R Core Team (2022). R: A language and environment for statistical computing. R Foundation for Statistical Computing, Vienna, Austria. <https://www.R-project.org/>

::: {.callout-tip title="Session Information" collapse="true"}
```{r}
#| label: session-info

utils::sessionInfo()
```
:::