---
title: "Simulation Suite"
subtitle: "v0.1 - The Digitial Sim"
author: "Dr Tom Moore - Statistical Scientist"
date: last-modified
output-file: "analysis_report.html"
format:
  html:
    toc: true
    toc-depth: 5
    toc-title: "Contents"
    theme: united
    highlight-style: tango
    fig-width: 9
    fig-height: 6
    code-tools: true
    code-fold: false
    code-overflow: scroll
    df-print: paged
    embeded-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| results: false

# Install packages and functions
if (!requireNamespace("pak", quietly = TRUE)) {
  install.packages("pak")
}

packages <- c(
  "tidyverse", "here", "janitor", "readxl", "modelr",
  "reshape2", "arrow", "pins", "usethis",
  "ggtext", "forcats", "cowplot", "ggdendro",
  "gplots", "GGally", "dagitty", "ggdag", "simDAG",
  "future", "furrr", "rstan", "bnlearn", "Rgraphviz",
  "brms", "bayesplot", "tidybayes", "cmdstanr", "purrr", "MASS",
  "nadiv", "graph4lg", "marginaleffects", "bayesplot"
)

# Install missing packages with pak
missing_pkgs <- packages[!sapply(packages, requireNamespace, quietly = TRUE)]
if (length(missing_pkgs) > 0) {
  pak::pkg_install(missing_pkgs, ask = FALSE)
  cat("Installed:", paste(missing_pkgs, collapse = ", "), "\n")
} else {
  cat("All Packages Installed :) \n")
}

# Load all packages
invisible(sapply(packages, library, character.only = TRUE))

bayesplot::color_scheme_set("blue")                                    # Set color scheme
options(mc.cores = parallel::detectCores())                            # Parallel Chains

# Stan Location: Production Sever Settings 
cmdstanr::set_cmdstan_path("/powerplant/workspace/hrltxm/workbench-k8s/stan/vanilla_stan/cmdstan-2.36.0")

# Import functions from source
source(here("Scripts/sim_functions.R"))
```

# Introduction

-  A minimal reproducible example of a simulation suite that demonstrates how environment, genetics, and experimental design play a role in recovering known population parameters and therefore implications for ecological inference. 


# Overview

The simulation should create an array (multiple matrices) that contains all the simulated trial information. Each matrix will represent a different variable (i.e., in this example "pid", "site", "a", "b", "x", "t1", "t2", "t3") that is created by unique function by different team members (with the intent that this will become a targets node): i.e.,


- Tom will create the functions to simulate environmental covariates (a and b) and the overall DAG (i.e., how all the matrices are related to one another based on the DGP or a DAG)

- Kathy will create the function to explore different trial designs + the number of checks to be included

- Poppy and Linley will create a function estimate the breeding values based on phylogeny

- Marcus will oversee the pipeline development (at the moment the pipeline isn't very flexible).


- At the end it would be good to have a function that iterates over the pipeline (i.e., similar to a power analysis - but would need to specify critorion to test).
- Should be flexible to different types of DAGS
- Should be parameterised in a way to showcase why it is important to consider experimental design, covariates, and genetic relationships when estimating breeding values.

# Simulate data
```{r}
#| label: simulated_data
#| fig.height: 6
#| fig.width:  12

# Create a 10*10 design matrix (n+1 = check)
init_design <- matrix(c(
  
  1, 2, 3, 4, 5,    1, 2, 3, 4, 5,
  6, 7, 8, 1, 2,    6, 7, 8, 1, 2,
  3, 4, 5, 6, 7,    3, 4, 5, 6, 7,
  8, 1, 2, 3, 4,    8, 1, 2, 3, 4,
  5, 6, 7, 8, 9,    5, 6, 7, 8, 9,
  
  1, 2, 3, 4, 5,    1, 2, 3, 4, 5,
  6, 7, 8, 1, 2,    6, 7, 8, 1, 2,
  3, 4, 5, 6, 7,    3, 4, 5, 6, 7,
  8, 1, 2, 3, 4,    8, 1, 2, 3, 4,
  5, 6, 7, 8, 9,    5, 6, 7, 8, 9

), nrow = 10, ncol = 10, byrow = TRUE)


# Create a 10*10 design matrix (n+1 = check)
bad_design <- matrix(c(
  
  1, 1, 1, 1, 1,    3, 3, 3, 3, 3,
  1, 1, 1, 1, 1,    3, 3, 3, 3, 3,
  1, 1, 9, 2, 2,    3, 3, 9, 4, 4,
  2, 2, 2, 2, 2,    4, 4, 4, 4, 4,
  2, 2, 2, 2, 2,    4, 4, 4, 4, 4,
  
  5, 5, 5, 5, 5,    7, 7, 7, 7, 7,
  5, 5, 5, 5, 5,    7, 7, 7, 7, 7,
  5, 5, 9, 6, 6,    7, 7, 9, 8, 8,
  6, 6, 6, 6, 6,    8, 8, 8, 8, 8,
  6, 6, 6, 6, 6,    8, 8, 8, 8, 8

), nrow = 10, ncol = 10, byrow = TRUE)


# Usage with default parameters:
sim <- sim_trial(seed = 123,          # Seed for reproducibility
                 n = 10,              # Trial size
                 n_sites = 4,         # Number of sites
                 eff_site = 10,       # Site effect size
                 
                 n_pid = 8,           # PID number
                 eff_pid = 1.2,       # PID effect size
                 n_chks = 4,          # Check number
                 per_chks = 0.90,     # Check percentile
                 
                 s_autocorr_a = 50,   # Climate autocorrelation
                 s_autocorr_b = 5000, # Soil autocorrelation
                 x_beta_a =  14.1,    # Climate effect on PID
                 x_beta_b = -15.7,    # Soil effect on PID
              
                 custom_pid_matrix = init_design # bad_design
                 )

sim$df_wide |> group_by(pid) |> tally()

print_matrices(sim$array, c("pid", "site"))

plot_sim_trial(sim, 
               plot_data = TRUE,
               standardize = TRUE,
               plot_dag = TRUE)
```

**Figure 1:** DAG of the simulated data that shows the relationships (arrows) between the outcome (Y), treatment (X), and other measured variables (A + B). The numbers on the arrows represent the known magnitude and direction of relationships with Y.
:::

::: {.panel-tabset}

## BLUPS

```{r}
#| label: blups
#| echo: false

# Models
mod_1 <- fit_brm(model_name = "mod_1", 
                 covariates = TRUE,
                 site_effects = TRUE)

mod_2 <- fit_brm(model_name = "mod_2", 
                 covariates = FALSE,
                 site_effects = TRUE)

mod_3 <- fit_brm(model_name = "mod_3", 
                 covariates = FALSE,
                 site_effects = FALSE)


plot_pid_effects(mod_1, resp = "t1")
plot_pid_effects(mod_2, resp = "t1")
plot_pid_effects(mod_3, resp = "t1")
```

- When there is no environmental information the number of checks is very important.

## Output

```{r}
#| label: model-output
#| echo: false
 
summary(mod_1)
summary(mod_2)
summary(mod_3)
```

## Validation

```{r}
#| label: validation
#| echo: false
#| message: false
#| fig-show: hold
#| results: false
#| layout-ncol: 3
#| fig.width: 4
#| fig.height: 2

pp_check(mod_1, resp = "t1")
pp_check(mod_2, resp = "t1")
pp_check(mod_3, resp = "t1")
```

:::

# References

McElreath, R. (2020). Statistical rethinking: A Bayesian course with examples in R and Stan. Chapman and Hall/CRC.

R Core Team (2022). R: A language and environment for statistical computing. R Foundation for Statistical Computing, Vienna, Austria. <https://www.R-project.org/>

::: {.callout-tip title="Session Information" collapse="true"}
```{r}
#| label: session-info

utils::sessionInfo()
``` 